---
title: "PS270 Final Project"
author: Margot Powell
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to congress, their political parties, the public's perception of the leaders, and the effectiveness of their time in office? Maybe something along the lines of "How does the party affiliation of members of Congress influence the number of bills they successfully sponsor and pass?".


I am still planning on researching the question "How does the party affiliation of members of Congress influence the number of bills they successfully sponsor and pass?". I am hoping to use the data source Legislative Effectiveness data set. I hypothesize that members in the majority party will create and pass more bills than those of the opposing party as a result of how Congress is structured. Institutionally, majority-party members benefit from leadership support and better access to legislative resources, which create systematic advantages. This then causes them to get reelected and only change when a new leadership gets in office. The main explanatory variable is party affiliation, specifically whether a member is part of the majority party or not. The outcome variable is the legislative success measured in the number of bills a member successfully sponsors that pass through. My hypothesis would be supported if the majority-party members, on average, pass significantly more bills then the minority party, and if the majority party stays in office longer then minority groups. It would be false if there is no difference in the number of bills passed between the two parties. 

```{r}
library(readxl)
congress <- read_excel("data/SLES_individual_through_2018.xlsx")
head(congress)
```

## First Graph 

```{r}
library(tidyverse)
laws_passed <- congress |>
  group_by(`1 = majority party`) |>
  summarize(
    avg_bills = mean(`Number of bills becoming law`, na.rm = TRUE)
  )
ggplot(laws_passed, aes(x = factor(`1 = majority party`), y = avg_bills, fill = factor(`1 = majority party`))) +
  geom_col()+
  labs(
    title = "Legislative Success by Majority-Party Status",
    x = "Party Status (Majority vs Minority)",
    y = "Average Number of Bills Becoming Law",
    fill = "Party Status"
  ) +
  scale_x_discrete(labels = c("0" = "Minority", "1" = "Majority")) +
  theme_minimal()
```
```{r}
library(broom)
model_1 <- lm(`Number of bills becoming law` ~ `1 = majority party`, data = congress)
knitr::kable(tidy(model_1))

```

### Write Up and Analysis

The figure shows that members of the majority party sponsor more bills that become law on average compared to members of the minority party. This pattern is consistent with my hypothesis that institutional advantages, such as agenda control and leadership support, increase legislative success for majority-party members. However, while the difference appears meaningful, further analysis using regression is necessary to determine whether this relationship is statistically significant.

The regression results show a strong relationship between majority-party status and legislative success. The coefficient on majority-party status is 2.847, which indicates that members of the majority party sponsor approximately 2.85 more bills that become law on average than members of the minority party. This suggests that majority-party membership is associated with substantially greater legislative success. This relationship is statistically significant (p < 0.001), meaning it is highly unlikely to have occurred by chance.


