---
title: "How Majority Status Shapes Lawmaking in Congress"
author: Margot Powell
description: "This project examines how majority-party status shapes legislative success in Congress, showing that the  institution of Congress plays a key role in determining which bills become law."
output:
  distill::distill_article:
    self_contained: false
---



## Introduction 
### Motivation
I am interested in exploring data related to Congress, including political parties, the public’s perception of leaders, and the effectiveness of their time in office. While these factors are often discussed separately, I am particularly interested in what actually translates into legislative success and how effective members of Congress are at getting things done. More specifically, this project is motivated by a broader question about how power operates within Congress. Political debates often focus on ideology or partisanship, but less attention is given to who is actually able to turn ideas into policy. By focusing on bill sponsorship and passage, this project aims to move beyond perception and capture a more concrete measure of political effectiveness.

### Hypothesis
This leads to the central research question: how does the party affiliation of members of Congress influence the number of bills they successfully sponsor and pass?

**I hypothesize that members in the majority party will create and pass more bills than those of the opposing party as a result of how Congress is structured.** Institutionally, majority-party members benefit from leadership support and better access to legislative resources, which create systematic advantages. These advantages increase the likelihood that their bills advance through the legislative process and ultimately become law.

## Data
I plan to use the Legislative Effectiveness dataset to analyze this relationship. This dataset allows me to compare members of Congress and evaluate their legislative output, making it well-suited for examining differences between majority and minority party members. The data is based on legislative activity across members of Congress.

The unit of analysis is the individual member of Congress in a given congressional term, meaning that each observation reflects a member’s legislative activity during a term in office. The research design is cross-sectional, as the analysis compares differences in legislative outcomes across members.

### Variables
The main **explanatory variable** is **party affiliation**, specifically whether a member is part of the majority party or not. 

The **outcome variable** is **legislative success**, measured by the number of bills a member sponsors that successfully move through the legislative process and become law. These outcomes are measured *per member per congressional term*, allowing for a consistent comparison across individuals.

### Criteria for Hypothesis
My hypothesis would be supported if majority-party members, on average, pass significantly more bills per term than minority-party members. It would not be supported if there is no meaningful difference in the number of bills passed between the two groups.

```{r,include = FALSE}
library(tidyverse)
library(readxl)
congress <- read_excel("data/SLES_individual_through_2018.xlsx")
head(congress)
```

## Results
### Distribution of Legislation Success

```{r,echo=FALSE}
ggplot(congress, aes(x = `Number of bills becoming law`)) +
  geom_histogram(binwidth = 1) +
  labs(
    title = "Distribution of Bills Becoming Law",
    x = "Number of Bills Becoming Law per Member per Term",
    y = "Count of Members"
  ) 
```
#### Write Up
The distribution of the number of bills becoming law per member per congressional term is highly skewed, with most members passing relatively few bills and a small number of members achieving higher levels of success. This suggests that legislative effectiveness is unevenly distributed across members of Congress, with a majority producing limited successful legislation while a smaller group accounts for a larger share of passed bills.

### Legislative Success by Majority-Party Status

```{r,echo=FALSE}
laws_passed <- congress |>
  group_by(`1 = majority party`) |>
  summarize(
    avg_bills = mean(`Number of bills becoming law`, na.rm = TRUE)
  )
ggplot(laws_passed, aes(x = factor(`1 = majority party`), y = avg_bills)) +
  geom_col(fill = c("darkblue", "steelblue"))+
  labs(
    title = "Legislative Success by Majority-Party Status",
    x = "Party Status (Majority vs Minority)",
    y = "Average Number of Bills Becoming Law\nper Member per Term",
    fill = "Party Status"
  ) +
  scale_x_discrete(labels = c("0" = "Minority", "1" = "Majority")) +
  theme_minimal()
```
#### Write Up
The figure shows that members of the majority party pass more bills, on average, per congressional term than members of the minority party. This pattern is consistent with my hypothesis that institutional advantages—such as agenda control, leadership support, and access to legislative resources—allow majority-party members to more effectively move legislation through Congress. While the difference appears meaningful, regression analysis is needed to determine the magnitude and statistical significance of this relationship.


### Regression

```{r, echo=FALSE}
library(broom)
library(modelsummary)
model_1 <- lm(`Number of bills becoming law` ~ `1 = majority party`, data = congress)
modelsummary(model_1)

```

### Regression Analysis
The regression results show a strong positive relationship between majority-party status and legislative success. The coefficient on majority-party status is approximately 2.85, which means that majority-party members pass about 2.85 more bills per member per congressional term than minority-party members. This is a meaningful difference and suggests that majority-party membership is associated with greater legislative effectiveness. Additionally, this means that majority-party members have a noticeable advantage in making legislative proposals into actual laws, proving the importance of your position in shaping policy outcomes.

The estimated effect is statistically significant (p < 0.001), indicating that the relationship is unlikely to be due to random chance. However, because this analysis is based on observational, cross-sectional data, the results should not be interpreted as  causal. Other factors such as seniority, committee assignments, or leadership positions may be having some influence on legislative success. 

## Conclusion
Overall, the results provide strong support for my hypothesis that members of the majority party are more legislatively effective than those in the minority party. Both the graphical analysis and regression results show that majority-party members pass more bills per congressional term, on average, than minority-party members. The regression coefficient indicates that majority-party members pass approximately 2.85 more bills per member per term, and this relationship is statistically significant, suggesting that the difference is not due to random chance. Given that most members pass relatively few bills per term, an increase of nearly three additional bills represents a substantively meaningful difference in legislative effectiveness.

These findings illustrate the idea that legislative success is not just a reflection of individual effort or ability, but is shaped by the institutional structures within Congress. Majority-party members benefit from control, support, and greater access to resources, all of which increase their ability to move legislation through. As a result, power within Congress plays a central role in determining who is able to translate ideas into actual policy outcomes.

However, this analysis has important limitations. Because the data is observational and cross-sectional, it is not possible to make strong causal claims about the effect of majority-party status on legislative success. Other factors, such as seniority or leadership positions may influence how many bills a member is able to pass. There could be a bias in this study.

Future research could improve this analysis by incorporating additional control variables, such as years in office, to better isolate the effect of majority-party status. Or, we could also explore whether this relationship differs between the House and Senate. Together, these additions would not only strengthen the analysis, but also deepen our understanding of how power operates within Congress. In conclusion, these findings suggest that legislative success in Congress is not only a function of individual effort, but is deeply shaped by institutional structures that determine who has the power to govern effectively but could also be strength throw more analysis.

```{r print-code, ref.label=knitr::all_labels(), echo = TRUE, eval=FALSE}

```

